### 2016-01-13 model/overflowmcmc.tex
Began writing the algorithm -- just the rejection/metropolis steps left. MCMC document updated on the fly to reflect specific algorithms used for e.g. truncated distributions

\subsection{Measurement error variance --- $\sigma^2$}
The full conditional distribution of the measurement error variance, $\sigma^2$, depends only on $\tilde{\bm{t}}_{1:n_{b}}$ and $\tau_{1:n_{b}}$. It is
\begin{align*}
p(\sigma^2|...) \propto \sigma^{-(n_{b} + \sum_{k=1}^{n_{b}}\tilde{n}_k)}\exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2\right]p(\sigma^2)
\end{align*}
where $p(\sigma^2)$ is the density of the prior distribution on $\sigma^2$. With the half-$t$ prior on $\sigma$ discussed above and letting $a_\sigma = (n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k)/2$ and $b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2$ the full conditional distribution of $(\sigma^2,\omega)$ is
\begin{align*}
p(\sigma^2,\omega|...) \propto (\sigma^2)^{-(a_\sigma+v)/2-1}\exp\left[-\frac{b_\sigma}{\sigma^2}\right]\exp\left[-\frac{\omega}{\sigma^2}\right]\omega^{(v + 1)/2 - 1}\exp\left[-\frac{\omega}{v s^2}\right].
\end{align*}
So we draw $\sigma^2$ and $\omega$ in two separate Gibbs steps:
\begin{enumerate}
\item Draw $\omega \sim G\left(\frac{v + 1}{2}, \frac{1}{\sigma^2} + \frac{1}{v s^2}\right)$ and form
\begin{align*}
a_\sigma = n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k &&\mbox{  and  }&& b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2/2;
\end{align*}
\item Draw $\sigma^2 \sim  IG\left(\frac{a_\sigma + v}{2}, b_\sigma + \omega\right)$.
\end{enumerate}

@@

\subsection{Transaction arrival times and sizes in MB --- $N(\tau_k)$ and $D_k$}
First consider transaction sizes, $\bm{d}=(d_1,d_2,\dots,d_{n_d})'$. Let $\bm{d}_k$ denote the sizes of the $M_k - M_{k-1}$ transactions that were stored in block $k$ so that $\bm{d} = (\bm{d}_1',\dots,\bm{d}_{n_{b}}')'$. The full conditional distribution of the $\bm{d}_k$'s is
\begin{align*}
p(\bm{d}_{1:n_{b}}|...)\propto \prod_{k=1}^{n_b}\prod_{i=1}^{M_k - M_{k-1}} d_i^{\alpha_d - 1}e^{-d_i\beta_d}1\left(\sum_{i=1}^{M_k - M_{k-1}} d_{ik}= D_k\right)1\left(d_{1k} > \bar{D}_{k-1} - D_{k-1} \mathrm{\ or\ } M_{k-1} = N_{k-1}\right).
\end{align*}
The second indicator function is required because if $M_{k-1} < N_{k-1}$ then $N_{k-1} - M_{k-1}$ transactionss must have overflowed into block $k$. So the amount of data from the first transaction of block $k$, $d_{1k}$, must have been greater than the difference between the amount of data in block $k-1$ and the data cap for block $k-1$. Similarly if $M_{k-1} = N_{k-1}$ then we know there was no overflow into block $k$, so we have no reason to think that $d_{1k}$ has a restricted support. Let $\bm{o}_k = \bm{d}_k/D_k$. Then the $\bm{o}_k$'s have potentially truncated independent Dirichlet distributions. That is
\begin{align*}
p(\bm{o}_{1:n_b}|...)\propto \prod_{k=1}^{n_b} \prod_{i=1}^{M_k - M_{k-1}}o_{ik}^{\alpha_d - 1} 1\left(d_{1k} > [\bar{D}_{k-1} - D_{k-1}]/D_{k} \mathrm{\ or\ } M_{k-1} = N_{k-1}\right).
\end{align*}
So the $\bm{d}$ step to draw from $p(\bm{d}|...)$ is as follows: Let $\bm{1}_n$ denote an $n$-vector of ones. For $k=1,2,\dots,n_b$
\begin{enumerate}
\item Draw $\bm{o}_{k} \sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1}})1\big(d_{1k} > 1[N_{k-1} - M_{k-1} > 0][\bar{D}_{k-1} - D_{k-1}]/D_{k}\big)$.
\item Set $\bm{d}_k = \bm{o}_kD_k$.
\end{enumerate}
When the dirichlet distribution is truncated the draw can be accomplished by rejection sampling, though other methods are available. [FILL IN WHICHEVER METHOD I END UP USING --- PROBABLY AUGMENTATION LIKE MY BLOG POST]

To draw the $N_k$'s first note that their support is restricted by the value of the $M_k$'s. Specifically for $k=1,2,\dots,n_b$ and $N_0 = 0$ we have $M_k \le N_k \le N_{k+1}$. Second, when 
\begin{align*}
\sum_{i=1}^{M_k - M_{k-1} + 1}d_{M_{k-1} + i} \le \bar{D}_k
\end{align*}
we know that $N_k = M_k$ because otherwise transaction $M_k + 1$ would be in block $k$. If the summation is greater than $D_k$ then it is still possible that $N_k = M_k$ because transaction $d_{M_k + 1}$ could still have arrived later than block $k$. In that case, the density of $N_k$ given $N_{0:(k-1)}$ and all other parameters and processes is given by
\begin{align*}
p(N_k|N_{0:(k-1)}, \dots) \propto \frac{\lambda_k^{N_k - N_{k-1}}e^{-\lambda_k}}{(N_k - N_{k-1})!}1(N_k \ge M_k)1(N_k \ge N_{k-1}).
\end{align*}
In other words, $\eta_k = N_k - N_{k-1}$ has a truncated Poisson distribution where the support is restricted to $\{M_k - N_{k-1}, M_k - N_{k-1} + 1, \dots \}$. [PUT INSTRUCTIONS FOR HOW TO DRAW FROM THE TRUNCATED POISSON HERE] So we can draw from $p(\bm{N}_{1:n_b}|\dots)$ as follows: For $k=1,2,\dots,n_b$
\begin{enumerate}
\item If $\sum_{i=1}^{M_k - M_{k-1} + 1} d_{M_k + i} \le \bar{D}_k$ set $N_k = M_k$.

>>>>

\subsection{Measurement error variance --- $\sigma^2$}
The full conditional distribution of the measurement error variance, $\sigma^2$, depends only on $\tilde{\bm{t}}_{1:n_{b}}$ and $\tau_{1:n_{b}}$. It is
\begin{align*}
p(\sigma^2|...) \propto \sigma^{-(n_{b} + \sum_{k=1}^{n_{b}}\tilde{n}_k)}\exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2\right]p(\sigma^2)
\end{align*}
where $p(\sigma^2)$ is the density of the prior distribution on $\sigma^2$. With the half-$t$ prior on $\sigma$ discussed above and letting $a_\sigma = (n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k)/2$ and $b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2$ the full conditional distribution of $(\sigma^2,\omega)$ is
\begin{align*}
p(\sigma^2,\omega|...) \propto (\sigma^2)^{-(a_\sigma+v)/2-1}\exp\left[-\frac{b_\sigma}{\sigma^2}\right]\exp\left[-\frac{\omega}{\sigma^2}\right]\omega^{(v + 1)/2 - 1}\exp\left[-\frac{\omega}{v s^2}\right].
\end{align*}
So we draw $\sigma^2$ and $\omega$ in two separate Gibbs steps:
\begin{enumerate}
\item Draw $\omega \sim G\left(\frac{v + 1}{2}, \frac{1}{\sigma^2} + \frac{1}{v s^2}\right)$ and form\footnote{The inconsistency in notation between the definition of $a_\sigma$ and $b_\sigma$ is a consequence of matching notation to the R code supplied for implementing this MCMC algorithm.}
\begin{align*}
a_\sigma = \frac{n_{b} + \sum_{k=1}^{n_b}\tilde{n}_k + v}{2} &&\mbox{  and  }&& b_\sigma = \sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k}(\tilde{t}_{ik} - \tau_k)^2;
\end{align*}
\item Draw $\sigma^2 \sim  IG\left(a_\sigma, b_\sigma/2 + \omega\right)$.
\end{enumerate}

@@

\subsection{Transaction arrival times and sizes in MB --- $N(\tau_k)$ and $D_k$}
First consider transaction sizes, $\bm{d}=(d_1,d_2,\dots,d_{n_d})'$. Let $\bm{d}_k$ denote the sizes of the $M_k - M_{k-1}$ transactions that were stored in block $k$ so that $\bm{d} = (\bm{d}_1',\dots,\bm{d}_{n_{b}}')'$. The full conditional distribution of the $\bm{d}_k$'s is
\begin{align*}
p(\bm{d}_{1:n_{b}}|...)\propto \prod_{k=1}^{n_b}\prod_{i=1}^{M_k - M_{k-1}} d_i^{\alpha_d - 1}e^{-d_i\beta_d}1\left(\sum_{i=1}^{M_k - M_{k-1}} d_{ik}= D_k\right)1\left(d_{1k} > \bar{D}_{k-1} - D_{k-1} \mathrm{\ or\ } M_{k-1} = N_{k-1}\right).
\end{align*}
The second indicator function is required because if $M_{k-1} < N_{k-1}$ then $N_{k-1} - M_{k-1}$ transactionss must have overflowed into block $k$. So the amount of data from the first transaction of block $k$, $d_{1k}$, must have been greater than the difference between the amount of data in block $k-1$ and the data cap for block $k-1$. Similarly if $M_{k-1} = N_{k-1}$ then we know there was no overflow into block $k$, so we have no reason to think that $d_{1k}$ has a restricted support. Let $\bm{o}_k = \bm{d}_k/D_k$. Then the $\bm{o}_k$'s have potentially truncated independent Dirichlet distributions. That is
\begin{align*}
p(\bm{o}_{1:n_b}|...)\propto \prod_{k=1}^{n_b} \prod_{i=1}^{M_k - M_{k-1}}o_{ik}^{\alpha_d - 1} 1\left(d_{1k} > [\bar{D}_{k-1} - D_{k-1}]/D_{k} \mathrm{\ or\ } M_{k-1} = N_{k-1}\right).
\end{align*}
So the $\bm{d}$ step to draw from $p(\bm{d}|...)$ is as follows: Let $\bm{1}_n$ denote an $n$-vector of ones. For $k=1,2,\dots,n_b$
\begin{enumerate}
\item If $N_k = M_k$ draw $\bm{o}_{k} \sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1}})$.

  Otherwise draw $\bm{o}_{k} \sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1}})1(d_{1k} > [\bar{D}_{k-1} - D_{k-1}]/D_k)$.
\item Set $\bm{d}_k = \bm{o}_kD_k$.
\end{enumerate}
To draw from the truncated Dirichlet distribution above, first note that by the properties of the Dirhichlet the marginal distribution of $o_{1k}$ is $Beta(\alpha_d, \alpha_d(M_k - M_{k-1} - 1))$ and conditional on $o_{1k}$, $\bm{o}_{-1,k}/(1 - o_{1k}) \sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1} - 1})$. Then $o_{1k}$ can be drawn from the truncated Beta distribution using the inverse cdf method, resulting in the following two step algorithm to draw from the desired truncated Dirichlet distribution:
\begin{enumerate}
\item Draw $u \sim U(0,1)$ and set $o_{1k} = F^{-1}(u*[1 - F((\bar{D}_{k-1} - D_{k-1})/D_k)])$ where $F$ is the cdf of the $Beta(\alpha_d, \alpha_d(M_k - M_{k-1} - 1))$ distribution.
\item Draw $\tilde{\bm{o}}_{2:(M_k - M_{k-1})}\sim Dir(\alpha_d\bm{1}_{M_k - M_{k-1} - 1})$ and set $o_{ik} = \tilde{o}_i$ for $i = 2,3,\dots,M_k - M_{k-1}$.
\end{enumerate}

To draw the $N_k$'s first note that their support is restricted by the value of the $M_k$'s. Specifically for $k=1,2,\dots,n_b$ and $N_0 = 0$ we have $M_k \le N_k \le N_{k+1}$. Second, when 
\begin{align*}
\sum_{i=1}^{M_k - M_{k-1} + 1}d_{M_{k-1} + i} \le \bar{D}_k
\end{align*}
we know that $N_k = M_k$ because otherwise transaction $M_k + 1$ would be in block $k$. If the summation is greater than $D_k$ then it is still possible that $N_k = M_k$ because transaction $d_{M_k + 1}$ could still have arrived later than block $k$. In that case, the density of $N_k$ given $N_{0:(k-1)}$ and all other parameters and processes is given by
\begin{align*}
p(N_k|N_{0:(k-1)}, \dots) \propto \frac{\lambda_k^{N_k - N_{k-1}}e^{-\lambda_k}}{(N_k - N_{k-1})!}1(N_k \ge M_k)1(N_k \ge N_{k-1}).
\end{align*}
In other words, $\eta_k = N_k - N_{k-1}$ has a truncated Poisson distribution where the support is restricted to $\{M_k - N_{k-1}, M_k - N_{k-1} + 1, \dots \}$. We can easily simulate from this distribution using the inverse cdf method. So we can draw from $p(\bm{N}_{1:n_b}|\dots)$ as follows: For $k=1,2,\dots,n_b$
\begin{enumerate}
\item If $\sum_{i=1}^{M_k - M_{k-1} + 1} d_{M_k + i} \le \bar{D}_k$ set $N_k = M_k$.
